The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics

### 2016-01-27 Papers/InProgress/IJCAI/submission.tex
Improve the paper at the sentence level

\title{An AI Success Story in Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument}
\author{Christoph Benzm\"uller\thanks{German Research Foundation DFG \ldots} and Bruno Woltzenlogel Paleo}
\author{}

@@

The in-depth computational analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similar to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL; cf. \cite{andrewsSEP} and the references
therein) even though G\"odel's proof is actually formulated in
higher-order \emph{modal} logic (HOML; cf. \cite{homl} and the
references therein). To bridge between the two logics we utilise and
further improve the logic embedding approach \cite{J23,C40}, which has
already been employed successfully in preceding related work.

The main novel contribution reported in this paper is a detailed
analysis (in various modal logics) of the inconsistency of G\"{o}del's
original version of the axioms used in his proof
\shortcite{GoedelNotes}. The extraction, reconstruction and
verification of an informal, human intuitive argument has been an open
problem since the first detection of this inconsistency by
Benzm\"uller and Woltzenlogel-Paleo \shortcite{C40} with the
\textsc{Leo-II} prover.  The verified refutation (discussed in
\S\ref{sec:inconsistency}) displays a surprisingly accessible
explanation of the inconsistency, which is philosophically profound
and never presented in the literature. The detection of this
inconsistency in combination with the work reported here thus
demonstrates that artificial intelligence systems -- particularly
higher-order automated theorem provers -- are capable of assisting in
the discovery and elucidation of \emph{new} and philosophically
relevant knowledge.

On the technical side, the quest for constructing a compelling
refutation, capable of convincing also human non-experts, led us to
improve the syntax of the embedding of modal logics in Isabelle/HOL
(as discussed in \S\ref{sec:improvedsyntax}). With the new syntax, a
(nearly) perfect match between the original pen and paper
presentations and our encoding in Isabelle/HOl is feasible. A more
user-friendly syntax, as reported here, is clearly an important
prerequisite for promoting the theorem proving technology employed
here to a wider community of philosophers, who are not necessarily
experts in automated reasoning or higher-order logic.

Another novel contribution reported here (in \S\ref{sec:improvedembedding}) is
the implementation of an alternative embedding for the only slightly
stronger modal logic \SFiveU. Our experiments have shown that the new
embedding is more efficient, as the following two, previously open problems can now be solved:
\begin{itemize}
\item Proving the final theorem T3 \textit{``Necessarily, there
    exists God''} (in Scott's \shortcite{ScottNotes} consistent variant) fully automatically and directly from the axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata).
\item Verifying automatically (in Isabelle/HOL) the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz \cite{ToDo:Leibniz}.  G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5}. The main difference to G\"odel's version is the
addition of a conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required. This difference has been
considered inessential and merely an oversight by many. For instance,
Hazen 1998, p.365 \cite[p.365]{Hazen} states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight--it is
included in related manuscripts''. However, the omitted conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  that G\"odel's axioms were inconsistent.}

@@

For more than four decades, the serious consequences of G\"odel's
mistake remained unnoticed, despite intense activity focused on
criticizing his argument. Especially since the discovery by Sobel
\cite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; each case needs to be analysed
  separately.}. Opponents of the argument
(e.g. Oppy \cite[p.226/227]{oppy96:_goedel_ontol_argum}
\cite[p.364]{oppy00:_respon_gettin}
\cite[p.1068]{oppy08:_higher_order_ontol_argum}) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

For a computational analysis of G\"odel's ontological argument in this
approach the embedding of HOMLs such as K,
KB and S5 with various domain conditions (possibilist and actualist)
is required. This idea has been successfully followed in related work
\cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, translated to predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HOML are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation
\cite{DBLP:journals/logcom/Ohlbach91}, which is here intra-logically
realized --- and extended for quantifiers --- in HOL by stating a set
of equations defining the logical constants (Fig.~\ref{QML_S5}). The
resulting object logic is the HOML K with rigid terms and constant
domains (possibilist quantifiers). Other logics (e.g. KB, S5) can be
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
means of an existence predicate, which can be used to guard the
quantifiers. Therefore, the embedding approach is very flexible.

\subsection{Improved Embedding}\label{sec:improvedembedding}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5}
\end{figure}

@@

The new embedding of \SFiveU in Isabelle is shown in Fig.~\ref{QML_S5}.
%
\marginpar{@Bruno: I commented your paragraph here, are BF and CBF
  equally valid in \SFiveU and \SFive?}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{``Necessarily, there exists God''}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

Wider adoption of higher-order theorem proving technology for reasoning about and within embedded object logics, especially among non-expert users, is still hindered by the gap between the syntax used by people, when they write logical formulas with pen and paper, and the syntax used by higher-order logic theorem provers. Even when the syntax of the underlying higher-order system is elegant (as is the case in Isabelle/HOL), the embedding of HOML into HOL may easily expose details of HOL that may be uncommon to the user, disturbing his/her experience while using the system. To illustrate this point, Fig.~\ref{UglyEssence} shows how the definition of essence looked like in previous work \cite{J28}, where advanced syntax-sugaring features were not used. It looks notably higher-order, and its style differs significantly from the common style seen in works on modal logics and the ontological argument. The following specific issues can be enumerated:
\begin{enumerate}
\item Lambda abstractions (which are a typical higher-order feature) appear explicitly in places where it didn't need to appear in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}).
\item Quantifiers appear as higher-order dqefined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the
  letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
  user, as it constantly reminds him/her that there is something
  unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he would expect, he has to read $(\psi y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

@@

In the typical workflow of trying to prove a conjecture with a theorem prover, it is customary to check the consistency of the axioms first. For if the axioms are inconsistent, anything (including the conjecture) would be trivially derivable in classical logic (\emph{ex falso quodlibet}). Surprisingly, when this routine check was performed on G\"odel's axioms \cite{C40}, the \textsc{Leo-II} prover claimed that the axioms were inconsistent. Unfortunately, the refutation generated by \textsc{Leo-II} was barely human-readable. The text file was 153 lines\footnote{Long lines with an average of 184 characters per line.} long and machine-oriented proof calculus (higher-order resolution \cite{W47}) and syntax (TPTP THF \cite{J22}) were used. Part of the file is displayed in Fig.~\ref{LEO-Proof}.

@@

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property $\lambda x. \bot$ (cf. Fig.~\ref{LEO-Proof}),
respectively, to self-difference $\lambda x.  x\not=x$.

Note that both terms have identical denotations in a logic setting
with full functional and Boolean extensionality as given
here. Nevertheless, some philosophers may actually prefer the use of
self-difference over the empty property\footnote{This occurred e.g. in
  a private communication of the authors with Andr\'e Fuhrmann.} in
the analysis below. However, for the proof to go through it is
irrelevant which notion we use and the reader may simply replace the
empty property by self-difference.

@@

\item To see that the sentence above is contradictory, we may reason semantically, in terms of possible worlds. If $w_0$ is the arbitrary current world, the $\pos$ operator forces the existence of a world $w$ accessible from $w_0$ such that $\nec \bot$ is true in $w$. But $\nec \bot$ can only be true in $w$, if there is no world $w'$ accessible from $w$. In logics with a reflexive or symmetric accessibility relation (e.g. \KB), it is easy to see that there must be a world $w'$ accessible from $w$: either $w'$ itself, in case of a reflexive relation, or $w_0$, in case of a symmetric relation. In fact, even in \K, with no accessibility condition, there must be a world $w'$ accessible from $w$. The reason is that $\pos \nec \bot$ should be \emph{valid} (true in all worlds). Therefore, it is true in $w$ as well, where the existence of an accessible world $w'$ is forced by the $\pos$ operator. As a model for $\pos \nec \bot$ (which is a consequence of G\"odel's axioms) cannot be built, G\"odel's axioms are inconsistent.
\end{enumerate}

Interestingly, the refutation automatically generated by
\textsc{Leo-II} uses a symmetric accessibility relation, and thus
requires the modal logic KB. The informal, human-constructed
refutation described above, on the other hand, requires only the
weaker modal logic K. In our experiments \textsc{Leo-II} (like all
other HOL provers) was still to weak to automatically prove the
inconsistency already in logic K. Hence, this remains an open problem for automated
theorem provers.

\subsection{Argument Reconstruction in Isabelle}

To verify the correctness of the informal argument explained above, it
was reconstructed in Isabelle/HOL, using Metis to automate the
inessential parts. The essential use of the Empty Essence Lemma, on
the other hand, is explicitly stated, to ensure that Isabelle is
reconstructing the same argument. In fact, without the help of this
lemma, Metis is not strong enough to refute G\"odel's axioms. 
\marginpar{ToDo: check this claim about Metis.}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Refutation of G\"odel's Axioms in Isabelle/HOl..} \label{InconsistencyIsabelleK}
\end{figure}

\subsection{Mapping the Inconsistency to G\"odel}
The inconsistency verified in Fig.~\ref{InconsistencyIsabelleK} follows from the definition of
essence (ess), the definition of necessary existence (NE), the
axioms A1(a) and A2 (which entail theorem T1), and axiom A5. It remains to demonstrate that
these ingredients are actually present in G\"odel's manuscript in
Fig.~\ref{GoedelScript}.

This can easily be seen: Axioms A1a in
Fig.~\ref{InconsistencyIsabelleK} is implied by Axiom Ax2 and the
highlighted footnote remark in Fig.~\ref{GoedelScript}. Axioms A2 and
A5 in Fig.~\ref{InconsistencyIsabelleK} correspond to Ax4 and Ax3 in
Fig.~\ref{GoedelScript}. The definitions of essence and necessary
existence are easy to identify. Thus, we conclude that the verified
inconsistency from Fig.~\ref{InconsistencyIsabelleK} does apply to 
G\"odel's original manuscript.

\section{Conclusion}\label{sec:conclusion}
The axioms and definitions in G\"odel's manuscript are inconsistent;
this has been detected automatically by the prover
\textsc{Leo-II}. Here we have presented a rational reconstruction and
verification of the inconsistency argument in Isabelle/HOL. This
argument is valid in all normal HOMLs including base logic K.
\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{./Images/Manuscript2.png}}
\caption{G\"{o}del's manuscript, with mutually inconsistent axioms and definitions highlighted.} \label{GoedelScript}
\end{figure*}

We have also presented several technical improvements regarding the
semantic embedding approach. In particular, we have a achieved a
nearly perfect match between pen and paper presentations in HOML and
the syntax in Isabelle/HOL. As a result, the embedding of HOML in HOL
is now fully transparent and, as we think, ready for wider adoption by
philosophers.

On the other hand, there is still room for improvements in
Isabelle/HOL. For example, Isabelle/HOL's Sledgehammer tool, which
orchestrates the calls to external automated theorem provers, such as
\textsc{Leo-II}, still fails to detect the inconsistency of G\"odel's
axioms, while a direct modeling of the problem in TPTP THF syntax in
combination with a direct call of \textsc{Leo-II} succeeds. In other
words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected. There are further
technical issues regarding Sledgehammer that motivate
improvements. For example, in default setting, Sledgehammer does not
immediately inform the user when a proof has been found and instead silently
 first executes a series of time-consuming proof analysis processes (e.g. it's
dependency minimizer), before it eventually informs the user of success. For the
involved proof problems presented here (e.g. G\"odel's theorem T3:
\textit{Necessarily, there exists God}), this phase of silence may
take several minutes (during which the user might actually give up on
the proof attempt), while \textsc{Leo-II} already reported success to
Sledgehammer much earlier (e.g. after 2,5 seconds for T3).

Moreover, our work points to a challenge for proof transformation:
even if \textsc{Leo-II}'s proof can eventually be automatically
verified in Isabelle, without a human-intuitive explanation of the
proof, this will presumably be of limited value for philosophers.  The
challenge is to extract from such a verification process an abstract
level argument, similar to what we have presented in this paper: for
philosophers, the reconstruction of such human-intuitive arguments
from machine found proofs will likely be crucial for accepting any
novel results.

@@

\section*{Acknowledgments}

Will be added at a later point.

>>>>

\title{The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics}
\author{Christoph Benzm\"uller\thanks{German Research Foundation DFG \ldots} and Bruno Woltzenlogel Paleo}
\author{}

@@

The in-depth computational analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similarly to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL; cf. \cite{andrewsSEP} and the references
therein), even though G\"odel's proof is actually formulated in
higher-order \emph{modal} logic (HOML; cf. \cite{homl} and the
references therein). To bridge between the two logics we utilise and
further improve the logic embedding approach \cite{J23,C40}, which has
already been employed successfully in preceding related work.

The main novel contribution reported in this paper is a detailed
analysis (in various modal logics) of the inconsistency of G\"{o}del's
original version of the axioms used in his manuscript
\shortcite{GoedelNotes}. The extraction, reconstruction and
verification of an informal, human intuitive argument has been an open
problem since the first detection of this inconsistency by
Benzm\"uller and Woltzenlogel-Paleo \shortcite{C40} with the
\textsc{Leo-II} prover.  The verified refutation (discussed in
\S\ref{sec:inconsistency}) displays a surprisingly accessible
explanation of the inconsistency, which is philosophically profound
and never presented in the literature. The detection of this
inconsistency in combination with the work reported here thus
demonstrates that artificial intelligence systems -- particularly
higher-order automated theorem provers -- are capable of assisting in
the discovery and elucidation of \emph{new} and philosophically
relevant knowledge.

On the technical side, the quest for constructing a compelling
refutation, capable of convincing also human non-experts, led to an
improvement of the syntax of the embedding of modal logics in Isabelle/HOL
(as discussed in \S\ref{sec:improvedsyntax}). With the new syntax, a
(nearly) perfect match between the original pen and paper
presentations and our encoding in Isabelle/HOL is feasible. A more
user-friendly syntax, as reported here, is clearly an important
prerequisite for promoting the theorem proving technology employed
here to a wider community of philosophers, who are not necessarily
experts in automated reasoning or higher-order logic.

Another novel contribution reported here (in \S\ref{sec:improvedembedding}) is
the implementation of an alternative embedding for the slightly
stronger modal logic \SFiveU. Our experiments have shown that the new
embedding is more efficient, as the following two, previously open problems can now be solved:
\begin{itemize}
\item Proving the final theorem T3 \textit{``Necessarily, there
    exists God''} (in Scott's \shortcite{ScottNotes} consistent variant) fully automatically and directly from the axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata). \marginpar{ToDo: 2004 is the year of Sobel's book, not of Scott's manuscript. We need to fix this as I fixed the date for Goedel's manuscript.}
\item Verifying automatically (in Isabelle/HOL) the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz. 
%\cite{ToDo:Leibniz}.  
G\"odel's manuscript
(Fig.~\ref{GoedelScript}(a)) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. The main difference to G\"odel's version is an extra conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}(a)) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required. This difference has been
considered inessential and merely an oversight by many. For instance,
Hazen [\citeyear[p.365]{Hazen}] states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
However, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5U}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  that G\"odel's axioms were inconsistent.}

@@

For more than four decades, the serious consequences of G\"odel's
mistake remained unnoticed, despite intense activity focused on
criticizing his argument. Especially since the discovery by Sobel
\shortcite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they to be analyzed
  separately.}. Opponents of the argument
(e.g. Oppy [\citeyear[p.226--227]{oppy96:_goedel_ontol_argum};
\citeyear[p.364]{oppy00:_respon_gettin};
\citeyear[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

For a computational analysis of G\"odel's ontological argument, the embedding of HOMLs such as K,
KB and S5 with various domain conditions (possibilist and actualist quantification)
is required. This idea has been successfully employed in related work
\cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, translated to predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HOML are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation
\cite{DBLP:journals/logcom/Ohlbach91}, which is here intra-logically
realized --- and extended for quantifiers --- in HOL by stating a set
of equations defining the logical constants (Fig.~\ref{QML_S5U}). The
resulting object logic is the HOML K with rigid terms and constant
domains (possibilist quantifiers). Other logics (e.g. KB, S5) can be
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
means of an existence predicate, which can be used to guard the
quantifiers. The embedding approach is, therefore, very flexible.

\subsection{Improved Embedding}\label{sec:improvedembedding}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5U.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5U}
\end{figure}

@@

The new embedding of \SFiveU in Isabelle is shown in
Fig.~\ref{QML_S5U}. It is important to note that we 
have  $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi$,
despite the differences in the underlying frame structures.

%
% \marginpar{@Bruno: I commented you paragraph here, since I am not
%   convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
%   and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{``Necessarily, there exists God''}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5U}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

Wider adoption of higher-order theorem proving technology for reasoning about and within embedded object logics, especially among non-expert users, is still hindered by the gap between the syntax used by people, when they write logical formulas with pen and paper, and the syntax used by higher-order logic theorem provers. Even when the syntax of the underlying higher-order system is elegant (as is the case in Isabelle/HOL), the embedding of HOML into HOL may easily expose details of HOL that may be uncommon to the user, disturbing his/her experience while using the system. To illustrate this point, Fig.~\ref{UglyEssence} shows how the definition of essence looked like in previous work \cite{J28}, where advanced syntax-sugaring features were not used. It looks notably higher-order, and its style differs significantly from the common style seen in works on modal logics and the ontological argument. The following specific issues can be enumerated:
\begin{enumerate}
\item Lambda abstractions (which are a typical higher-order feature) appear explicitly in places where they didn't need to appear in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}(a)).
\item Quantifiers appear as higher-order defined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the
  letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
  user, as it constantly reminds him/her that there is something
  unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he would expect, he has to read $(\psi ~ y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

@@

In the typical workflow during an attempt to prove a conjecture with a theorem prover, it is customary to check the consistency of the axioms first. For if the axioms are inconsistent, anything (including the conjecture) would be trivially derivable in classical logic (\emph{ex falso quodlibet}). Surprisingly, when this routine check was performed on G\"odel's axioms \cite{C40}, the \textsc{Leo-II} prover claimed that the axioms were inconsistent. Unfortunately, the refutation generated by \textsc{Leo-II} was barely human-readable. The text file was 153 lines\footnote{Long lines with an average of 184 characters per line.} long and used a machine-oriented proof calculus (higher-order resolution \cite{W47}) and syntax (TPTP THF \cite{J22}). Part of the file is displayed in Fig.~\ref{LEO-Proof}.

@@

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property $\lambda x. \bot$ (cf. Fig.~\ref{LEO-Proof}).
%
Note that the terms for the empty property ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in a logic setting
with full functional and Boolean extensionality as given
here. Nevertheless, some philosophers may actually prefer the use of
self-difference over the empty property\footnote{This occurred in
  a private communication of the authors with Andr\'e Fuhrmann.} in
the analysis below. However, for the proof to go through it is
irrelevant which notion we use and the reader may simply replace the
empty property by self-difference. In HOL it is customary to think of a set as a unary predicate that is true for (and only for) elements of the set. Therefore, the empty property (i.e. the unary predicate that is false for any individual) represents the empty set, and hence we use the empty set symbol ($\emptyset$) as an abbreviation for the empty property. \marginpar{ToDo: should we use the empty set symbol in the isabelle code in figure 5 as well?}

@@

\item To see that the sentence above is contradictory, we may reason semantically, thinking of possible worlds. If $w_0$ is the arbitrary current world, the $\pos$ operator forces the existence of a world $w$ accessible from $w_0$ such that $\nec \bot$ is true in $w$. But $\nec \bot$ can only be true in $w$, if there is no world $w'$ accessible from $w$. In logics with a reflexive or symmetric accessibility relation (e.g. \KB), it is easy to see that there must be a world $w'$ accessible from $w$: either $w'$ itself, in case of a reflexive relation, or $w_0$, in case of a symmetric relation. In fact, even in \K, with no accessibility condition, there must be a world $w'$ accessible from $w$. The reason is that $\pos \nec \bot$ should be \emph{valid} (true in all worlds). Therefore, it is true in $w$ as well, where the existence of an accessible world $w'$ is forced by the $\pos$ operator. As a model for $\pos \nec \bot$ (which is a consequence of G\"odel's axioms) cannot be built, G\"odel's axioms are inconsistent.
\end{enumerate}

Interestingly, the refutation automatically generated by
\textsc{Leo-II} uses a symmetric accessibility relation, and thus
requires the modal logic KB. The informal, human-constructed
refutation described above, on the other hand, requires only the
weaker modal logic K. In our experiments \textsc{Leo-II} (like all
other HOL provers) was still too weak to automatically prove the
inconsistency already in logic K. Hence, this remains an open problem for automated
theorem provers.

\subsection{Argument Reconstruction in Isabelle}  \label{sec:arg2}
To verify the correctness of the informal argument explained above, it
was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
  external provers such as \textsc{Leo-II} or Satallax, 
  constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
inessential parts. The essential use of the Empty Essence Lemma, on
the other hand, is explicitly stated, to ensure that Isabelle is
reconstructing the same argument. In fact, without the help of this
lemma, Metis is still not strong enough to refute G\"odel's
axioms. \marginpar{ToDo: check this claim about Metis.}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Refutation of G\"odel's Axioms in Isabelle/HOL.} \label{InconsistencyIsabelleK}
\end{figure}

\subsection{Mapping the Inconsistency to G\"odel}
The inconsistency verified in Fig.~\ref{InconsistencyIsabelleK} follows from the definition of
essence (ess), the definition of necessary existence (NE), the
axioms A1(a) and A2 (which entail theorem T1), and axiom A5. It remains to show that
these ingredients are actually present in G\"odel's manuscript in
Fig.~\ref{GoedelScript}(a).

This can easily be seen: Axioms A1a in
Fig.~\ref{InconsistencyIsabelleK} is implied by Axiom Ax2 and the
highlighted footnote remark in Fig.~\ref{GoedelScript}(a). Axioms A2 and
A5 in Fig.~\ref{InconsistencyIsabelleK} correspond to Ax4 and Ax3 in
Fig.~\ref{GoedelScript}(a). The definitions of essence and necessary
existence are easy to identify. Thus, we conclude that the verified
inconsistency from Fig.~\ref{InconsistencyIsabelleK} does apply to 
G\"odel's original manuscript.

\subsection{Inconsistency of G\"odel's Axioms in \SFiveU}

Isabelle/HOL's Sledgehammer tool, which orchestrates the calls to
external automated theorem provers, such as \textsc{Leo-II}, still
fails to detect the inconsistency of G\"odel's axioms in the standard
embedding, while a direct modeling of the problem in TPTP THF syntax
in combination with a direct call of \textsc{Leo-II} succeeds. In
other words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected.
\begin{figure*}
\centerline{
\includegraphics[width=.65\textwidth]{./Images/Manuscript2.png} \hfill
\includegraphics[width=.35\textwidth]{./Images/Inconsistency_S5U_direct.png}
}
\caption{(a) G\"{o}del's manuscript, with mutually inconsistent axioms and
  definitions highlighted; (b) Refutation in
\SFiveU} \label{Inconsistency_S5U} \label{GoedelScript} 
\end{figure*}

\marginpar{ToDo: Are we really supposed to use \$ instead of ``Sec.''?}
On the other hand (and further confirming the claims from \$\ref{sec:improvedembedding}), 
the reconstruction in Isabelle/HOL with the improved embedding for \SFiveU was more efficient: 
the inconsistency could be detected
by \textsc{Leo-II} also when called via
Sledgehammer. Moreover, the result could subsequently be verified with
Metis even without the Empty Essence Lemma; cf. Fig.\ref{Inconsistency_S5U}. 
Interestingly, Metis' argument is different from that of
\$\ref{sec:arg1} and \$\ref{sec:arg2}, since axiom A1b is not used.

\section{Conclusion}\label{sec:conclusion}
The axioms and definitions in G\"odel's manuscript are inconsistent;
this has been detected automatically by the prover
\textsc{Leo-II}. Here we have presented a rational reconstruction and
verification of the inconsistency argument in Isabelle/HOL. This
argument is valid in all normal HOMLs including base logic K.

We have also presented several technical improvements regarding the
semantic embedding approach. In particular, we have a achieved a
nearly perfect match between pen and paper presentations in HOML and
the syntax in Isabelle/HOL. As a result, the embedding of HOML in HOL
is now fully transparent and ready for wider adoption by
philosophers.

On the other hand, there is still room for many pragmatical
improvements in Isabelle; just one example: in default setting,
Sledgehammer does not immediately inform the user when a proof has
been found and instead silently first executes a series of
time-consuming proof analysis processes (e.g. its dependency
minimization), before it eventually reports success. For G\"odel's
theorem T3 (\textit{Necessarily, there exists God}), for example, this
phase of silence takes several minutes --- during which the user might
actually give up on the proof attempt --- even though \textsc{Leo-II} already
reported success to Sledgehammer after 2,5 seconds.

More importantly, our work reveals a challenge for automated reasoning:
the (so far partially manual) extraction of an informal argument from a formal proof. 
Without accompanying human-understandable explanations,
the proofs generated by provers such as \textsc{Leo-II} or Metis, will
presumably be only of limited value for philosophers, for whom intuitive
arguments remain crucial for the acceptance of novel results.

@@

%\section*{Acknowledgments} Will be added at a later point.
%\paragraph*{Acknowledgments} Will be added at a later point.

%\pagebreak
